---
title: "credit.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
bibliography: '../../notes/epsy8264.bib'
csl: '../../notes/style/apa-single-spaced.csl' 
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print = "75")
opts_chunk$set(
  echo = FALSE,
  prompt = FALSE,
  tidy = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE
  )
opts_knit$set(width = 75)
```


The data in [credit.csv](https://github.com/zief0002/epsy-8264/raw/master/data/credit.csv) contains simulated credit card data for 400 individuals. These data are from @James:2013. The variables in the data set are:

- `balance`: Customer's average credit card balance (in dollars)
- `income`: Customer's reported income (in \$10,000 dollars)
- `limit`: Credit limit issued to customer
- `rating`: Customer's credit rating; higher values indicate a better credit rating
- `cards`: Number of credit cards the customer has
- `age`: Customer's age
- `education`: Number of years of education


### Preview

```{r}
credit = readr::read_csv(file = "https://github.com/zief0002/epsy-8264/raw/master/data/credit.csv")
print(head(credit), width = Inf)
```


### References
