---
title: "iowa-judges.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


The data in [contraception.csv](https://github.com/zief0002/epsy-8264/raw/master/data/contraception.csv) were collected from several sources (e.g., World Bank). The variables are:

- `country`: Country name
- `region`: Region of the world
- `contraceptive`: Percentage of women who are practicing, or whose sexual partners are practicing, any form of contraception. It is usually measured for women ages 15&ndash;49 who are married or in union.
- `educ_female`: Average number of years of formal education (schooling) for females
- `gni`: Categorical measure of the economy indicating if the country has a low or high gross national income




### Preview

```{r}
contraception = readr::read_csv(file = "~/Documents/github/epsy-8264/data/contraception.csv")
head(contraception)
```


### References


Roser, M. (2017). [Fertility rate](https://ourworldindata.org/fertility-rate#empowerment-of-women). *Our world in data*.

UNICEF. (2016). [State of the world's children 2016](https://www.unicef.org/publications/files/UNICEF_SOWC_2016.pdf). *United Nations Population Division's World Contraceptive Use, household surveys including Demographic and Health Surveys and Multiple Indicator Cluster Surveys.*

World Bank (2019). [World Bank open data](https://data.worldbank.org/). 

