---
title: "mcycle.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
bibliography: '../../notes/epsy8264.bib'
csl: '../../notes/style/apa-single-spaced.csl'    
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


This data, stored in [mcycle.csv](https://github.com/zief0002/epsy-8251/raw/master/data/keith-gpa.csv) contains data from a series of simulated motorcycle accidents during an experiment on the efficacy of crash helmets. Since, helmets are designed to help mediate the amount of acceleration experienced by the head, it is important to understand how well they actually do this after a crash. To study this researchers took accelerometer readings of a test dummy's head after it crashed. The attributes in this data set are:

- `times`: Time after impact (in milliseconds)
- `accel`: Head acceleration (in g)



### Preview

```{r}
mcycle = readr::read_csv(file = "~/Documents/github/epsy-8264/data/mcycle.csv")
print(head(mcycle), width = Inf)
```


### References

<!-- <p style="padding-left: 50px; text-indent: -50px;"> -->

<p style="padding-left: 50px; text-indent: -50px;">Silverman, B. W. (1985) Some aspects of the spline smoothing approach to non-parametric curve fitting. *Journal of the Royal Statistical Society series B, 47*, 1--52.</p>

<p style="padding-left: 50px; text-indent: -50px;">Venables, W. N., \& Ripley, B. D. (1999) *Modern applied statistics with S-PLUS* (3rd ed.). New York: Springer.</p>
