---
title: "stack-1979.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
bibliography: '../../notes/epsy8264.bib'
csl: '../../notes/style/apa-single-spaced.csl'
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
#library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


@Stack:1979 studied predictors of income inequality in a paper published in the *American Sociological Review*. He posited that more political participation and a strong Socialist party in a country would be associated with less income inequality. To control for variation in economic development, he included a measure of energy consumption, which had been identified as a reasonable economic proxy in previous studies. Stack's data, stored in [stack-1979.csv](https://raw.githubusercontent.com/zief0002/epsy-8264/master/data/stack-1979.csv), include four attributes measured on 18 countries. The attributes are:

- `country`: Country name
- `inequality`: Ratio of the share of income received by the most wealthy population quintile (richest 20\%) to the share received by the poorest 40\% of the population; Higher values indicate more income inequality
- `turnout`: Proportion of the adult population voting in the most recent national election prior to 1972
- `energy`: Energy consumption per capita (expressed in million metric tons of coal equivalents; higher values indicate more economic development
- `socialist`: Annual average proportion of seats held by socialist parties in the national legislature, over the first twenty postwar years

### Preview

```{r}
options(dplyr.width = Inf, width = 85)
stack = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-8264/master/data/stack-1979.csv")
head(stack)
```


### References
