---
title: "states-2019.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
bibliography: '../../notes/epsy8264.bib'
csl: '../../notes/style/apa-single-spaced.csl'  
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print = "75")
opts_chunk$set(
  echo = FALSE,
  prompt = FALSE,
  tidy = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE
  )
opts_knit$set(width = 75)
```


The data in [states-2019.csv](https://github.com/zief0002/epsy-8264/raw/master/data/states-2019.csv) include statistics collected from Wikipedia, the 2019 American Community Survey, and the National Centers for Environmental Information. The attributes in the data are:

- `state`: State/territory name
- `life_expectancy`: Life expectancy (in years)
- `population`: Population estimate (in millions)
- `income`: Per capita income (in thousands of dollars)
- `illiteracy`: Illiteracy rate (in percent)
- `murder`: Murder and non-negligent manslaughter rate (per 100,000 population)
- `hs_grad`: Percentage of high school graduates
- `frost`: Mean number of days with minimum temperature below freezing
- `area`: Land area (in square miles)




### Preview

```{r}
usa = readr::read_csv(file = "~/Documents/github/epsy-8264/data/states-2019.csv")
print(head(usa), width = Inf)
```


### References




