---
title: "Matrix Computations in R"
subtitle: "EPsy 8264"
author: Andrew Zieffler"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, servr.daemon = TRUE)
```


# Enter a Matrix

.pull-left[
$$\mathbf{X} = \begin{bmatrix}
1 & -2 & 3 \\ 
4 & -5 & -6 \\ 
7 & 8 & 9 \\ 
0 & 0 & 10
\end{bmatrix}$$

To enter a matrix in R, use the `matrix()` function. The elements of the matrix will be filled-in by columns.
]

.pull-right[

```{r}
X = matrix(
  data = c(1,  4, 7,  0, 
          -2, -5, 8,  0, 
           3, -6, 9, 10),
  nrow = 4,
  ncol = 3
  )

X
```
]

---

# Fill Elements By Row

.pull-left[
The `byrow=TRUE` argument will fill the elements by rows rather than columns.
]

.pull-right[
```{r}
# Enter elements by row
X = matrix(
  data = c(1, -2,  3,
           4, -5, -6,
           7,  8,  9,
           0,  0, 10),
  byrow = TRUE, #<<
  nrow = 4
  )

X
```
]

---

# Your Turn

Enter the matrix **B** into R.

$$\mathbf{B} = \begin{bmatrix}
-5 & 1 & 3 \\ 
2 & 2 & 6 \\ 
7 & 3 & -4
\end{bmatrix}$$

--

```{r}
B = matrix(
  data = c(-5, 1,  3,
            2, 2,  6,
            7, 3, -4),
  byrow = TRUE,
  nrow = 3
  )
```


---

# Vectors

.pull-left[
A matrix with a single column is referred to as a **column vector**. 

$$\mathbf{a} = \begin{bmatrix}
2 \\ 
0 \\ 
1 \\
3
\end{bmatrix}$$

]

.pull-right[
```{r}
# Enter column vector
a = matrix(
  data = c(2, 0, 1, 3),
  byrow = TRUE,
  nrow = 4
  )

a
```
]

---

.pull-left[
A matrix with a single row is referred to as a **row vector**. 

$$\mathbf{b} = \begin{bmatrix}
-1 & 6 & 0 & 9 \end{bmatrix}$$

]

.pull-right[
```{r}
# Enter row vector
b = matrix(
  data = c(-1, 6, 0, 9),
  byrow = TRUE,
  nrow = 1
  )

b
```
]


---

# Dimensions of a Matrix

.pull-left[
$$\underset{4\times 3}{\mathbf{X}} = \begin{bmatrix}
1 & -2 & 3 \\ 
4 & -5 & -6 \\ 
7 & 8 & 9 \\ 
0 & 0 & 10
\end{bmatrix}$$

$$\underset{4\times 1}{\mathbf{a}} = \begin{bmatrix}
2 \\ 
0 \\ 
1 \\
3
\end{bmatrix}$$
$$\underset{1\times 4}{\mathbf{b}} = \begin{bmatrix}
-1 & 6 & 0 & 9 \end{bmatrix}$$
]

.pull-right[
The `dim()` function will return the dimensions of a matrix.

```{r}
# Dimensions of X
dim(X)

# Dimensions of column vector a
dim(a)

# Dimensions of row vector b
dim(b)
```
]
---

# Your Turn

What are the dimensions of **B**?

$$\mathbf{B} = \begin{bmatrix}
-5 & 1 & 3 \\ 
2 & 2 & 6 \\ 
7 & 3 & -4
\end{bmatrix}$$

Use R to verify the dimensions.

--

```{r}
dim(B)
```

---

# Transpose

.pull-left[
The `t()` function will produce the transpose of a matrix.

$$\begin{split}\mathbf{b} &= \begin{bmatrix}
-1 & 6 & 0 & 9 \end{bmatrix}\\[1em]
\mathbf{b^{\prime}} &= \begin{bmatrix}
-1 \\ 6 \\ 0 \\ 9 \end{bmatrix}\end{split}$$
]

.pull-right[
```{r}
# Matrix transpose
t(b)

# Dimensions of transpose
dim(t(b))
```

]

---

.pull-left[
$$\begin{split}\mathbf{X} &= \begin{bmatrix}
1 & -2 & 3 \\ 
4 & -5 & -6 \\ 
7 & 8 & 9 \\ 
0 & 0 & 10
\end{bmatrix}\\[3em]
\mathbf{X^{\prime}} &= \begin{bmatrix}
1 & 4 & 7 & 0 \\ -2 & -5 & 8 & 0 \\ 3 & -6 & 9 & 0 \end{bmatrix}\end{split}$$
]

.pull-right[
```{r}
# Matrix transpose
t(X)

# Dimensions of transpose
dim(t(X))
```

]

---

# Your Turn

Find the transpose of $\mathbf{B}$ and the dimensions of $\mathbf{B}^{\prime}$.

$$\mathbf{B} = \begin{bmatrix}
-5 & 1 & 3 \\ 
2 & 2 & 6 \\ 
7 & 3 & -4
\end{bmatrix}$$

Use R to verify.

--

.pull-left[
```{r}
# Transpose of B
t(B)
```
]

.pull-right[
```{r}
# Dimensions
dim(t(B))
```
]

---

# Adding/Subtracting Matrices

.pull-left[
Matrices that have the same dimensions can be added/subtracted.

$$\underset{2\times 3}{\mathbf{D}} = \begin{bmatrix}
1 & 2 & 3 \\ 
4 & 5 & 6
\end{bmatrix}$$

$$\underset{2\times 3}{\mathbf{E}} = \begin{bmatrix}
-5 & 1 & 2 \\ 
3 & 0 & 4
\end{bmatrix}$$

The resulting matrix has the same dimensions as the originals.
]

.pull-right[
We add/subtract elements in the same position.

$$\begin{split}\mathbf{D}+\mathbf{E} &= \begin{bmatrix}
1+-5 & 2+1 & 3+2 \\ 
4+3 & 5+0 & 6+4
\end{bmatrix} \\[1em]
&= \begin{bmatrix}
-4 & 3 & 5 \\ 
7 & 5 & 10 \end{bmatrix}\end{split}$$

$$\begin{split}\mathbf{D}-\mathbf{E} &= \begin{bmatrix}
1--5 & 2-1 & 3-2 \\ 
4-3 & 5-0 & 6-4
\end{bmatrix} \\[1em]
&= \begin{bmatrix}
6 & 1 & 1 \\ 
1 & 5 & 2 \end{bmatrix}\end{split}$$
]

---


.pull-left[
```{r}
# Input D
D = matrix(
  data = c(1, 2,  3,
           4, 5,  6),
  byrow = TRUE, 
  nrow = 2
  )

# Input E
E = matrix(
  data = c(-5, 1, 2,
            3, 0, 4),
  byrow = TRUE, 
  nrow = 2
  )
```
]

.pull-right[
```{r}
# Add matrices
D + E

# Subtract matrices
D - E
```
]


---

# Multiplication by a Scalar

.pull-left[
A scalar is a $1\times1$ matrix (a number). A matrix and a scalar can be multiplied together by multiplying each element in the matrix by the scalar.

$$\begin{split} 3 \times \mathbf{D} &= 3\begin{bmatrix}
1 & 2 & 3 \\ 
4 & 5 & 6
\end{bmatrix}\\[1em]
&= \begin{bmatrix}
3\times1 & 3\times2 & 3\times3 \\ 
3\times4 & 3\times5 & 3\times6
\end{bmatrix} \\[1em]
&= \begin{bmatrix}
3 & 6 & 9 \\ 
12 & 15 & 18
\end{bmatrix}\end{split}$$
]

.pull-right[
```{r}
# Multiplication by a scalar
3 * D
```

]

---

# Diagonal Elements

.pull-left[
The `diag()` function will return the diagonal elements of a square matrix.

$$\mathbf{B} = \begin{bmatrix}
-5 & 1 & 3 \\ 
2 & 2 & 6 \\ 
7 & 3 & -4
\end{bmatrix}$$
]

.pull-right[
```{r}
diag(B)
```
]

---

# Your Turn

Find the diagonal for $\mathbf{B}^{\prime}$ (the transpose of $\mathbf{B}$).

$$\mathbf{B} = \begin{bmatrix}
-5 & 1 & 3 \\
2 & 2 & 6 \\
7 & 3 & -4
\end{bmatrix}$$

Use R to verify.

--

```{r}
# Diagonal of B'
diag(t(B))
```

---

# Diagonal of Non-Square Matrices

.pull-left[
The `diag()` function also works on non-square matrices. However, it returns the elements on the diagonal starting with the element in the $[1,1]$ position.

$$\mathbf{X} = \begin{bmatrix}
1 & -2 & 3 \\ 
4 & -5 & -6 \\ 
7 & 8 & 9 \\ 
0 & 0 & 10
\end{bmatrix}$$
]

.pull-right[
```{r}
diag(X)
```
]

---

# Matrix Trace

.pull-left[
The trace of a matrix is the sum of its diagonal elements

$$\mathbf{B} = \begin{bmatrix}
-5 & 1 & 3 \\ 
2 & 2 & 6 \\ 
7 & 3 & -4
\end{bmatrix}$$
]

.pull-right[
```{r}
# Compute trace of B
sum(diag(B))
```
]


You can also use the `tr()` function from the **psych** library to compute the trace.

```{r}
# Compute trace of B
psych::tr(B)
```

---

# Your Turn

Find the trace for $\mathbf{B}^{\prime}$ (the transpose of $\mathbf{B}$).

$$\mathbf{B} = \begin{bmatrix}
-5 & 1 & 3 \\
2 & 2 & 6 \\
7 & 3 & -4
\end{bmatrix}$$

Use R to verify.

--

```{r}
# Diagonal of B'
sum(diag(t(B)))
```

---

# Identity Matrix

.pull-left[
The `diag()` function can also be used to create an *identity matrix*. The argument is the number of rows and columns.

$$\underset{3\times 3}{\mathbf{I}} = \begin{bmatrix}
1 & 0 & 0 \\ 
0 & 1 & 0 \\ 
0 & 0 & 1
\end{bmatrix}$$
]

.pull-right[
```{r}
# Create a 3x3 identity matrix
diag(3)
```
]

